---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
set.seed(42)
```

# Overview regression models

General comments:

-   [Introduction to linear regression models](https://theoreticalecology.github.io/AdvancedRegressionModels/2A-LinearRegression.html)
-   what is the distribution of your response?
    -   Normally distribution –\> linear regression `lm()` @sec-lm and [here](https://theoreticalecology.github.io/AdvancedRegressionModels/2A-LinearRegression.html)
    -   Binary response (0, 1, 0, ...) -\> logistic regression `glm(..., family = binomial())` @sec-logistic and [here](https://theoreticalecology.github.io/AdvancedRegressionModels/4A-GLMs.html#or-kn-data---logistic-regression)
    -   Count data -\> Poisson glm `glm(.., family = poisson())` @sec-poisson and [here](https://theoreticalecology.github.io/AdvancedRegressionModels/4A-GLMs.html#count-data---poisson-regression)
-   Syntax:
    -   Addictive effects `y~a+b`
    -   Interactions `y~a*b`, **scale your variables first!!!** For more details on how to interpret interactions see [here](https://theoreticalecology.github.io/AdvancedRegressionModels/2A-LinearRegression.html#interactions) and [why is it so important so scale/center variabls](https://theoreticalecology.github.io/AdvancedRegressionModels/2A-LinearRegression.html#centering-and-scaling-of-predictor-variables)
    -   Quadratic effects `y~a+I(b^2)` **scale your variables first!!!**
-   Visualize effects via the `library(effects)` package
-   Interpretation: take the link/inverse link function in account when interpreting glms (see @sec-glm for more details)
-   Check residuals, however:
    -   for `lm`, you can just plot the model afterwards `plot(model)`
    -   for `glm`, you have to use the `DHARMa` package `simulateResiduals(model, plot=TRUE)`

::: column-margin
Variables can be scaled with the `scale(...)` function: `df$height <- scale(df$height)`
:::

::: column-margin
Install the `effects` package via `install.packages("effects")`
:::

## Linear regression

Normally distributed response:

```{r}
data(airquality)
str(airquality)
summary(airquality)
pairs(airquality)

plot(Ozone ~ Temp, data = airquality)

fit <- lm(Ozone ~ Temp, data = airquality)
abline(fit, col = "red")
summary(fit)
```

```{r}
library(effects)
plot(allEffects(fit))
plot(allEffects(fit, partial.residuals = T))

pairs(airquality)
fit <- lm(Ozone ~ Temp + Wind + Solar.R , data = airquality)
summary(fit)
plot(allEffects(fit, partial.residuals = T))

# optional scale to standardize effect sizes 
# scale command by default divides by standard deviation and 
# subracts the mean
fit <- lm(Ozone ~ scale(Temp) + scale(Wind) + scale(Solar.R), data = airquality)
summary(fit)

# centering is NOT OPTIONAL = you have to centering if using numeric
# variables with interactions (important thing is to center)

fit <- lm(Ozone ~ Temp * Wind , data = airquality)
summary(fit)
plot(allEffects(fit, partial.residuals = T))
# main effects change when changing * to +

fit <- lm(Ozone ~ scale(Temp) * scale(Wind)  , data = airquality)
summary(fit)
# main effects do not change when changing * to +
# in this case, can interpret main effects as the average 
# effect (e.g. of Temp, Wind) in the range of the data 

# residual checks
fit <- lm(Ozone ~ scale(Temp) + scale(Wind)  , data = airquality)
summary(fit)
plot(allEffects(fit, partial.residuals = T))

par(mfrow = c(2,2))
plot(fit)

# doesn't really look optimal, maybe a bit of nonlinearity
# and residuals don't look normal

fit <- lm(sqrt(Ozone) ~ scale(Temp) * scale(Wind)  , data = airquality)
summary(fit)

par(mfrow = c(2,2))
plot(fit)

# summary(aov(fit))
# too difficult for you to interpret probably, better stay with 
# effect sizes

# categorical variables 
par(mfrow = c(1,1))

boxplot(weight ~ group, data = PlantGrowth, notch = T)
fit <- lm(weight ~ group, data = PlantGrowth)
summary(fit)

summary(aov(fit))
summary(aov(weight ~ group, data = PlantGrowth))
```

## Logistic regression

```{r}
library(carData)
data(TitanicSurvival)
str(TitanicSurvival)

TitanicReduced = TitanicSurvival[complete.cases(TitanicSurvival), ]

fit <- glm(survived ~ age + sex + passengerClass , family = binomial, 
           data = TitanicReduced)

summary(fit)

plot(allEffects(fit))

library(DHARMa)
res <- simulateResiduals(fit)
plot(res)

# not perfect, let's see variables against predictor
plotResiduals(res, TitanicReduced$age)
plotResiduals(res, TitanicReduced$passengerClass)

# nothing to find here - maybe play around with this yourself 
# and see if you find the solution - as a hint: when
# including interactions between the variables, you will
# find significant interactions and a nicely fitting model

# if you have k/n data (e.g. k of n people in the same group survived), 
# you would specify it like this

# fit <- glm(cbind(survived, notSurvived) ~ pred , family = binomial, data = myData)
# fit <- glm(survived ~ pred , family = binomial, data = myData, weight = totalTrials)
```

## Poisson regression

```{r}
library(glmmTMB)

data("Owls")
str(Owls)

fit <- glm(SiblingNegotiation ~ SexParent + offset(log(BroodSize)) , 
           family= poisson, data = Owls)

summary(fit)

# note: in Poisson GLMs, offset(log(time)) standardizes to observation time / area

library(DHARMa)
res <- simulateResiduals(fit)
plot(res)

# Discussion how to improve the fit for this model in https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html#owl-example-count-data
```

## Multinomial regression

For multinomial regression, see page 69 of our Essential Statistics lecture notes https://www.dropbox.com/s/fxozlnzd5ntfntk/EssentialStatistics.pdf?dl=0
