---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Basic data operations with R

## Data manipulation in R

R works like a calculator:

```{r}
2+2
5*4
2^2
```

Always pay attention to the parenthese to set the order of the calculations:

```{r}
2*4^3-1
2*4^(3-1)
(2*4)^3-1
(2*4)^(3-1)
```

We can also use **functions** that perform specific calculations:

```{r}
sqrt(4)
sum(c(2,2))
abs(-1)
```

We can assign values/data to **objects**:

```{r}
object.name <- 1
variable1 <- c(2,2) # c() function to concatenate/group values
```

Note that both operators '\<-' or "=" work when creating a object.

Functions in R (e.g. `sum()`, `mean()`, etc.) have arguments that control/change their behavior and are also used to pass the data to the function:

```{r}
mean(x = c(2, 2))
mean(variable1)
```

::: {.callout-tip appearance="default"}
## Help

A list and description of all arguments in a function can be found in the help of a function (which can be accessed via `?mean` or `help(mean)`, or if you place the cursor on the function and press F1 in Rstudio).
:::

### Data types and data structures

There are four important data types in R (there are more but we focus on these 5):

-   **Numeric:** 1, 2, 3, 4

-   **Logical**: TRUE or FALSE

-   **Characters**: "A", "B",...

-   **Factors**: "A", "B",... which are characters but we have to tell R explicitly that they are factors - have levels

-   Not Available/Not a Number: **NA, NaN** (empty value)

Based on the data types we can build data structures which contain either only specific data types or a mixture of data types:

-   **Vector**: Several values of **one** data type, can be created with the `c` function:

    ```{r}
   num.vec <- c(5, 3, 5, 6) # numeric vector
   log.vec <- c(TRUE, TRUE, FALSE, TRUE) # logical vector
   char.vec <- c("A", "B", "C") # character vector
   factor.vec <- as.factor(c("A", "B", "C")) # factor vector
    ```

-   **Matrix**: two-dimensional data structure of **one** data type, can be created with the `matrix` function (we can pass a vector to the matrix function and tell it via arguments how the matrix should be constructed):

    ```{r}
    mat1 <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3, ncol = 2)
    ```

-   **Data.frame**: Often our data has variables of different types which makes a matrix unsuitable data structure. Data.frames can handle different data types and is organized in columns (one column = one variables) and can be created with the `data.frame` function:

    ```{r}
    data1 <- data.frame(A = c(1, 2, 3), B = c("A", "B", "C"), 
                        C = c(TRUE, FALSE, FALSE))
    ```

- **List**: A list is a data structure that can contain elements of different types and different lengths, like a bag where you can include anything. It is created with the `list` function:

```{r}
    my.list <- list(A = c(1, 2, 3), # a vector
                    B = data1,      # a data.frame
                    C = mat1)      # a matrix
```



### Data manipulation

A vector is a one dimensional data structure and we can access the values by using `[ ]`:

```{r}
vec = c(1, 2, 3, 4, 5)
vec[1] # access first element
vec[5] # access last element
```

A data.frame is a two dimensional data structure. Let's define a data.frame from two vectors:

```{r}
df = data.frame(
  x = c(2, 2, 2, 3, 2, 2, 1), # add column named x
  y = c(4, 5, 5, 4, 5, 3, 5)  # add a second column named y
)
#Let's see how this looks like:
df
```

Access parts of the data.frame:

```{r}
df[1,2] # get element in row 1, column 1
df[7,1] # get element in row 7, column 1
df[2,]  # get row 2
df[,2]  # get column 2
# or use the $ sign to access columns:
df$y
df[2:4,1:2] # get rows 2 to 4 and only columns 1 and 2
```

We can also set filters:

```{r}
df[df$x > 2, ] # show only data where x is larger than 2
df[df$y == 5, ] #show only data where y equals 5
df[df$y == 5 & df$x == 1, ] #show only data where y equals 5 AND x equals 1
df[df$y == 5 | df$x == 3, ] #show data where y equals 5 OR x equals 3
```

::: {.callout-tip appearance="default" collapse="true"}
### Logical operators

| Operators | Meaning               |
|-----------|-----------------------|
| \<        | Lower than             |
| \<=       | Lower than or equal to |
| \>        | Higher than             |
| \>=       | Higher than or equal to |
| ==        | Equal to              |
| !=        | Not equal to          |
| !a        | Not a                 |
| a\|b      | a or b                |
| a & b     | a and b               |
| isTRUE(a) | Test if a is true     |

: Logical operators in R
:::

Add an additional column with NA values:

```{r}
df$NAs = NA # fills up a new column named NAs with all NA values
df
```

## Data analysis workflow

This is a simple version of what you're going to learn during this course:

1.  Let's say we measured the size of individuals in two different treatment groups

    ```{r}
    group1 = c(2, 2, 2, 3, 2, 2, 1.1)
    group2 = c(4, 5, 5, 4, 5, 3, 5.1) 

    class(group2)
    ```

2.  Descriptive statistics and visualization

    ```{r}
    mean(group1)
    mean(group2)

    boxplot(group1, group2)
    ```

3.  Testing for differences. Question: Is there a difference between group1 and group2?

    ```{r}
    t.test(group1, group2)
    ```

4.  Interpretation of the results. Individuals in Group 2 were larger than those in group 1 (t test, t = -6.62, p \< 0.0001)

In the course we will work a lot with datasets implemented in R or in R packages which can be accessed via their name:

```{r}
dat = airquality
head(dat)
```
