---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
set.seed(42)
```

# Exercise - GLM {-}

In this exercise you will practice to:

-   choose the correct model for your data
-   interpret the model output of generalized linear models
-   check the residuals using the DHARMa package

Before you start, remember to clean your global environment (if you haven't already) using `rm(list=ls())`.

For your analyses, you will use the datasets `nitrofen` and `melanoma` which are both implemented in the `boot` package. You will also need the packages `DHARMa` and `effects`. Please install and load these packages, if you haven't done this yet:

```{r, eval=T}
library(boot)
library(DHARMa)
library(effects)
```

## Analyzing the `nitrofen` dataset

The Ministry of Agriculture has appointed your university to investigate the toxicity of the herbicide nitrofen on the fertility of the waterflea species `Ceriodaphnia dubia` in order to asses the implications of the herbicide for ecosystems.\
In an experiment conducted by your fellow researchers, the offspring of the waterflea species `Ceriodaphnia dubia` were counted as a function of different concentrations of the herbicide. Your job is to do the analysis.

::: {.callout-warning}
### Question

Carry out the following tasks:

-   Convert the variable `conc` into a factor.
-   Fit a suitable model for the relationship between total amount of offspring (`total`) and nitrofen concentration (`conc`).
-   Test for overdispersion.
-   Interpret and plot the effect.
-   Test for the overall effect of nitrofen concentration.

You may need the following functions:

-   `str()`
-   `as.factor()`
-   `glm()`
-   `summary()`
-   `testDispersion()`
-   `plot()`
-   `allEffects()`
-   `anova(..., test = "Chisq")`

Use your results to answer the following questions:

**1.You analyzed the response of Ceriodaphnia dubia to different concentrations of the herbicide nitrofen. Looking at your results, which of the following statements is correct? (Several anwers are correct)**

```{r}
#| include: false
opts_p <- c(
  answer= "The MLE for the mean number of offspring at a nitrofen concentration of 0 is 31.4 (Don't forget to apply the (inverse) link function to the estimate).",
  "At a nitrofen concentration of 160, Ceriodaphnia dubia produced significantly less offspring than in the control treatment (conc = 0).",
  answer = "At a nitrofen concentration of 235, Ceriodaphnia dubia produced significantly less offspring than in the control treatment (conc = 0).",
  answer = "The residuals show significantly less variance than would be expected from a Poisson distribution (underdispersion)."
)
```
`r longmcq(opts_p)`

**2. How do you test for the overall significance of concentration on the total amount of offspring? Perform the respective test to check whether the variable conc as a whole has an effect on the total amount of offspring. How many asteriks are displayed for significance?**

```{r}
#| include: false
opts_p <- c(
  "none = not significant",
  "* =  p < 0.05",
  "** =  p < 0.01",
  answer = "*** =  p < 0.001"
)
```
`r longmcq(opts_p)`


:::

`r hide("Click here to see the solution")`

```{r nitrofen}
# prepare data
nitrofen$conc <- as.factor(nitrofen$conc) # change variable conc to a factor

# plot the relation that we want to fit
plot(total ~ conc, nitrofen)

# Fit a suitable model for the relationship between total amount of offspring (total) and nitrofen concentration (conc):
fit <- glm(total ~ conc, family = "poisson", data=nitrofen)
summary(fit)

# test for overdispersion
testDispersion(fit)

# plot effect
plot(allEffects(fit))

# log link to calculate predicted values at the response scale:
# predicted response = exp(Intercept + Estimate * predictor)
exp(3.44681) # or
exp(coef(fit)[1])

# Test for the overall effect of *conc* on the total number of offspring
anova(fit, test = "Chisq")
```

`r unhide()`

## Analyzing the `melanoma` dataset

In the following, you will investigate the `melanoma` dataset provided by the University Hospital of Odense, Denmark. The data are of patients with malignant melanomas. You can find a more detailed description in the help of `melanoma`. The question you want to investigate is whether the occurrence of ulceration (`ulcer`, yes = 1, no = 0) is related to the thickness of the tumor (`thickness` in mm) and the sex of the patient (`sex`, male = 1, female = 0) and the interaction of the two.

::: {.callout-warning}
### Question

Carry out the following tasks:

-   Fit an appropriate model to answer the research question.
-   Check the model residuals.
-   Interpret and plot all effects including an analysis of the deviance.

You may need the following functions:

-   `str()`
-   `glm()`
-   `summary()`
-   `anova()`
-   `simulateResiduals()`
-   `plot()`
-   `allEffects()`

Use your results to answer the following questions:

**1. Let's move to the melanoma dataset: Assuming you chose the correct distribution, which type of regression did you use for the analysis?**


```{r}
#| include: false
opts_p <- c(
  "Linear regression",
  answer = "Logistic regression",
  "Poisson regression"
)
```
`r longmcq(opts_p)`

**3. You have checked the residuals of your model. Which of the following patterns and conclusions apply? (Several answers are correct)**

```{r}
#| include: false
opts_p <- c(
  answer = "There remains a significant pattern between the residuals and the fitted values.",
  answer = "An additional quadratic term for thickness could help to improve the model.",
  "The residuals deviate from the expected distribution.",
  "The model is overdispersed.",
  answer = "The model assumptions are not met yet, and I don't trust this model without further improvements."
)
```
`r longmcq(opts_p)`


:::

`r hide("Click here to see the solution")`

- *There remains a significant pattern between the residuals and the fitted values.* --> You can see this in the right part of the DHARMa residual plot.
- *An additional quadratic term for thickness could help to improve the model.* --> Try it out!
- *The residuals deviate from the expected distribution.* --> You can see this in the left part of the DHARMa residual plot.
- *The model is overdispersed.* --> Only poisson or binomial k/n models can be overdispersed!


```{r melanoma}
# get the data
dat <- melanoma

# Fit an appropriate model to answer the research question.
dat$sex <- as.factor(dat$sex) # change variable sex to factor (this is optional, not necessary with binary values 0/1)
fit <- glm(ulcer ~ thickness * sex, family = "binomial", data=dat)

# Check residuals
res <- simulateResiduals(fit, n = 500)
plot(res)

# model interpretation
summary(fit)
anova(fit, test = "Chisq")

# Calculate Pseudo R2: 1 - Residual deviance / Null deviance
1 - 234.04/281.13 # Pseudo R2 of model

# plot effects
plot(allEffects(fit))
```

As the residuals look quite suspicious and all quantile regressions significantly deviate, we can try to improve the model with a quadratic term for thickness.

```{r}
fit <- glm(ulcer ~ thickness * sex + I(thickness^2), family = "binomial", data=dat)
summary(fit)
res <- simulateResiduals(fit)
plot(res)

# plot effects
plot(allEffects(fit))
```

The quadratic term solves the problem of the residuals. The effects plots look quite different. There seems to be a maximum of ulcer around a thickness of 10 that we would have missed without the quadratic term.
`r unhide()`


## Analyzing Elk data

Download the elk_data from GRIPS and import the file using the `load( )` function.

:::{.column-margin}
RData is a R specific data type. You can save any R object by running `save(object, file "filename.RData")`
:::

::: {.callout-warning}
#### Example - Elk Data

You will be given a data set of habitat use of Elks in Canada. Measured is the presence of Elks (0/1), and a number of other predictors. Description of variables:

-   `dist_roads` - distance of the location to the next road

-   `NDVI` - normalized difference vegetation index, essentially greeness of vegetation on the site

-   `ruggedness` of the terrain

-   `dem` - digital eleveation model = elevation above sea level

-   `presence` - presence of the elk

-   `habitat` - open or forest

Perform either:

a)  A predictive analysis, i.e. a model to predict where Elks can be found.
b)  A causal analysis, trying to understand the effect of roads on Elk presence.

**Hints**:

- Hypothesis: presence ~ dist_roads
- What are potential confounders? How can you see if a variable is a confounder?

:::

`r hide("Click here to see the solution")`
A. Predictive analysis

```{r}
load(file = "elk_data.RData")

library(MASS)
fit <- glm(presence ~ dist_roads  + dem + ruggedness, data = elk_data, family = "binomial")
predictive_model = MASS::stepAIC(fit, direction = "both")
summary(predictive_model)

```

B. Causal analysis

The predictive model has actually dropped the variable of interest (distance to roads) which shows the risks of tools that select for the best predictive model such as AIC selection: Collinear variables that we need to adjust our effects, are often dropped.

For the causal model, we really need to think about the causal relationships between the variables:

We are interested in the effect of dist_roads on presence:

```{r}
summary(glm(presence ~ dist_roads, data = elk_data, family = "binomial"))
```

Positive effect of dist_roads on elk, or in other words, more elks closer to the roads? Does that make sense? No, we expect a negative effect!

Altitude (dem) and the `ruggedness` probably affect both variables, `presence` and `dist_roads`, and thus they should be considered as confounders:

```{r}
fit = glm(presence ~ dist_roads+ dem + ruggedness, data = elk_data, family = "binomial")
```

The effect of `dist_roads` is now negative!

Let's check the residuals:

```{r}
library(DHARMa)
res <- simulateResiduals(fit, plot = TRUE)
plot(res, quantreg = TRUE)
plotResiduals(res, form = elk_data$dem, quantreg = TRUE)
plotResiduals(res, form = elk_data$ruggedness, quantreg = TRUE)
```

The functional forms of our confounders are not perfect.

Since we are not really interested in them, a cool trick is to use a GAM (generalized addictive model) which automatically adjusts the functional for of the fitted curve to flexibly take care of the confounders. Our main predictor `dist_roads` is still modelled as a linear effect.

```{r}
library(mgcv)
fit2 <- gam(presence ~ dist_roads + s(dem) + s(ruggedness), data = elk_data, family = "binomial")
summary(fit2)
plot(fit2)
```

Let's take another look at the residual plots, in particular for the confounders.

```{r}
res <- simulateResiduals(fit2, plot = TRUE)
plot(res, quantreg = TRUE)
plotResiduals(res, form = elk_data$dem, quantreg = TRUE)
plotResiduals(res, form = elk_data$ruggedness, quantreg = TRUE)
```

Now, everything looks perfect
`r unhide()`
