---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
set.seed(42)
```

# Exercise - Multivariate statistics {.unnumbered}

```{r load data and libraries, echo=FALSE, message=FALSE, warning=FALSE}
dat = read.table(file = "http://evol.bio.lmu.de/_statgen/Multivariate/11SS/RIKZGroups.txt", header = T)
library(vegan)
library(cluster)
library(ape)
library(ggfortify)
```

In this exercise you will practice analyzing multivariate datasets using:

-   ordination methods which explore higher-order correlations among variables and similarities between observations
-   clustering methods which aggregate individual observations into clusters.

### Useful functions for multivariate data analyses

`pairs()` - create pairplot (plot all variables against each other)\
`prcomp()` - calculate PCA\
`plot(pca_object)` - plot variance explained by PC axes\
`biplot()` - plot PCA\
`metaMDS()` - calculate NMDS\
`ordiplot(nmds_object)` - plot NMDS\
`dist()` - create distance matrix\
`hclust()` - perform hierarchical clustering; apply to distance matrix\
`plot(as.phylo(hc_object))` - plot dendrogram from hclust output\
`kmeans()` - perform kmeans non-hierarchical clustering; access cluster assignments using `kmeans_object\$cluster`

### Background on the dataset

Marine coastal ecosystems harbor the most productive and diverse communities on Earth. However, they are extremely vulnerable to climate change and human activities (such as landclaim, pollution, recreation purposes, ...) As a consequence, the performance of this ecosystem has decreased considerably. To better understand these ecosystems, the dutch governmental institute (RIKZ: Rijksinstituut voor Kust en Zee) started a research project on sandy beaches to investigate the relationship between environmental factors and how they might affect benthic fauna. *(from the LMU and Zuur, Ieno, Smith (2007), Chapter 12.8-12.10 and 27)*

**Question**\
We now want to use ordination methods to explore

-   the variability in environmental conditions
-   the variability in species richness

We also want to use clustering methods to define 3 environmental types and hierarchically cluster the samples with respect to their species richness.

**Dataset**

Read in the dataset as follows:

```{r, eval = FALSE}
dat = read.table(file = "http://evol.bio.lmu.de/_statgen/Multivariate/11SS/RIKZGroups.txt", header = T)
head(dat)
```

You already know the functions `str()` and `summary()` to get an overview of a dataset and to see which variables the dataset contains.

-   columns 2:5 is **species richness** within four larger faunal groups
-   columns 7:16 are different **environmental variables**

Not important for us are:

-   column 6 (week), a time stamp\
-   column 17 (sorting1), variable from the observational design

Let's get into the analysis!

## Unconstrained ordination (PCA)

Carry out the following analyses:

-   Make two PCAs, one for the environmental and one for the species richness data (see columns above).
-   Name the results `pca_environment` and `pca_species`.
-   Create a biplot for each PCA.
-   Create a barplot for the proportion of variance explained by each component.

For example, the result for species richness should look like this:

```{r, echo = FALSE}
pca_species <- prcomp(dat[,2:5], scale = T) # conduct a principle component analysis (pca) with the assinged name "pca"

par(mfrow = c(1, 2))
biplot(pca_species, main = "PCA") # plot the results of the pca as a rotation matrix
barplot(summary(pca_species)$importance[2,1:4],
        main = "Components of species",
        ylab = "Proportion of variance explained")
par(mfrow = c(1, 1))
```

You need the following functions and the package `vegan`:

-   `prcomp()`
-   `biplot()`\
-   `barplot()`
-   `summary()`

```{r, eval = FALSE}
library(vegan)
```

**Hints**

-   Don't forget to scale the variables in the PCA.
-   In order to get the proportion of the explained variance, have a look at the summary of your analysis. `str()` shows you what the summary contains. You want to specifically look at the `importance`, here the second row contains the proportion of variance of all components. So what we want to plot is: `summary(pca_species)$importance[2,]`.
-   Don't forget to give your plots a title.

Now, use your results to **answer the following questions:**

::: callout-warning
### 1. Task

**1. In which multivariate dataset do the first and second components explain more variance?**

```{r}
#| include: false
opts_p <- c( "Both explain the same amount of variance in the first two compontents",
             "Environment",
             answer = "Species")
```

`r longmcq(opts_p)`

`r hide("Click here to see the solution")`

```{r}
## 1.) Conduct the principle component analyses (PCAs):

# a) PCA of environmental data
pca_environment = prcomp(dat[,7:16], scale = T)
summary(pca_environment)

op <- par(mfrow = c(1,2))
biplot(pca_environment, main = "PCA")            # plot the results of the PCA as a rotation matrix
barplot(summary(pca_environment)$importance[2,], # get the importance measure
     main = "Components of environment",
     ylab = "Proportion of variance explained")

# b) PCA of species richness data
pca_species = prcomp(dat[,2:5], scale = T)
summary(pca_species)

biplot(pca_species, main = "PCA")            # plot the results of the PCA as a rotation matrix
barplot(summary(pca_species)$importance[2,], # get the importance measure
     main = "Components of species",
     ylab = "Proportion of variance explained")
par(op)
```

From the *summary()* output we can see that the first and second components explain more variance in the species PCA.

`r unhide()`
:::

::: callout-warning
### 2. Task

2.  What are the two samples that have the highest score on the respective third PCA axes?

`r hide("Click here to see the solution")`

```{r, eva=F}
# -> order the samples by their PC3 coordinate:
order(pca_environment$x[,'PC3'], decreasing = T) # -> 24 is highest
order(pca_species$x[,'PC3'], decreasing = T) # -> 7 is highest
```

`r unhide()`
:::

::: callout-warning
### 3. Task

3.  From looking at the biplot: Which environmental variables would you choose, if you have only the resource to measure three variables?

```{r}
#| include: false
opts_p <- c(
   answer = "salinity, penetrability and grain size",
   "NAP, angle 1 and chalk",
   "grain size, chalck, exposure")
```

`r longmcq(opts_p)`

`r hide("Click here to see the solution")`

```{r}
biplot(pca_environment);
abline(v=0, lty="dashed");abline(h=0, lty="dashed")
```

From looking at the biplot, we choose 3 variables that describe a lot of variation (i.e. have a large length in the biplot) and have little correlation with others. An appropriate choice would be salinity, penetrability and grain size. `r unhide()`
:::

::: callout-warning
### 4. Task

4.  Describe the following correlations:
    a)  salinity and humus
    b)  angle1 and exposure
    c)  Polychaeta and Mollusca

`r hide("Click here to see the solution")`

We can get information on the correlations of variables by looking at their representation in the biplot:

a)  salinity and humus: same direction -\> positively correlated

b)  angle1 and exposure: opposite directions -\> negatively correlated

c)  Polychaeta and Mollusca: almost perpendicular -\> uncorrelated

`r unhide()`
:::

## Clustering

### K-means

We want to use clustering methods to define 3 environmental types. Use the function `kmeans()` with `centers = 3` (number of clusters to be generated).

::: callout-caution
Remember to set a seed: the choice of the first k centers is random.
:::

```{r, eval = FALSE}
set.seed(467456)
cl = #...
```

Compare the three clusters with the result of the PCA using the following code that uses another plotting framework called ggplot2:

```{r, eval = FALSE}
library(ggfortify)
autoplot(pca_environment, colour = cl$cluster, size = 2, 
         loadings = TRUE, loadings.colour = 'black',
         loadings.label = TRUE, loadings.label.size = 3,
         loadings.label.colour = "black")
```

The colors of the points represent the three clusters. 

**Answer the following questions**:

::: callout-warning
### 1. Task
-   How is it possible that four observations in the middle (in red - if you have used the same seed) belong to a different cluster than the observations around them (in black)?

`r hide("Click here to see the solution")`

```{r}
# Create clusters using all environmental variables
set.seed(467456)
cl <- kmeans(dat[,7:16], centers = 3) # 3 clusters to be generated

# Plot
library(ggfortify)
autoplot(pca_environment, colour = cl$cluster, size = 2, 
         loadings = TRUE, loadings.colour = 'black',
         loadings.label = TRUE, loadings.label.size = 3,
         loadings.label.colour = "black")
```

To understand why the four observations in the middle belong to a different cluster than the observations around them, we must take into account, that the biplot is only a 2-dimensional representation of a more-than-2 dimensional space. Therefore, the four points in the middle will be dissimilar to the points around them with respect to a variable that is not well represented by the first two PCA axes.

`r unhide()`
:::

::: callout-warning
### 2. Task
-   Which environmental variables are on average high within the black cluster (cluster 1)?

`r hide("Click here to see the solution")`
Environmental variables that are on average high within cluster 1:
looking at the plot we find that exposure and grain size are high on average within cluster 1
`r unhide()`
:::


### Hierarchical clustering

Now we want to hierarchically cluster the samples with respect to their species richness, as shown in the following plot:

```{r clustering, echo = FALSE, fig.height=10}
hc = hclust(dist(dat[, 2:5]))
plot(as.phylo(hc), tip.color = dat$week)
```

Create the same plot using the functions:

-   `hclust()`\
-   `plot()`
-   `as.phylo()`

Load the package `ape`. Have a look at the help for `hclust()` to read what the function does and look at the examples for further help on how to use the function. Then have a look at what the function `as.phylo()` does. Now, color the labels using the variable `week`. You can do this using the argument "tip.color =".

::: callout-warning
### 1. Task
Which of the following statements are correct? (could be more than 1)

Hint: To be able to read the plot more easily, you can click *Zoom* in the top pane of the Plots window in RStudio.

```{r}
#| include: false
opts_p <- c( 
  answer = "Observations 2 and 25 are more similar than 2 and 20.",
  answer = "The observations taken during Week 1 (here in black) have a more similar species richness than those taken during Week 2 (here in red).",
  "The tree cleary identifies 4 distinct groups of species diversity.",
  "Observations 6 and 35 are as similar as 7 and 8.")
```
`r longmcq(opts_p)`

`r hide("Click here to see the solution")`

```{r solution-HierarClcust}
library(ape)

hc = hclust(dist(dat[, 2:5]))
plot(as.phylo(hc), tip.color = dat$week)
```

`r unhide()`
:::
