---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
set.seed(42)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
treedata <- read.csv('./data/treedata.csv', sep = ';')
species <- read.csv('./data/species.csv', sep = ';')
```

# Data handling

## Exercise

In this exercise you will practice common steps of the data preparation procedure using a dataset that was gathered in a forest in Switzerland.

The main steps are as follows (**to solve the tasks, please carefully read the detailed instructions and hints below**):

1.  Read the provided datasets *treedata.csv* and *species.csv* to R.
2.  Have a look at the dataset properties.\
3.  Find values that prevent a column you expect to be numeric to do so.\
4.  Does the dataset contain NA-values in `height`? How many?\
5.  Have a look at the data: Check for implausible values. For now, remove these values.\
6.  Add the species names to the dataset.\
7.  Create a new dataset containing only trees with both, height and dbh measurements.\
8.  Are there any correlations within the new dataset?\
9.  Remove all trees with `rem` = F4 from the dataset\
10. Calculate mean dbh by species.

------------------------------------------------------------------------

To thorougly check the dataset and perform the operations, you will need the following functions:

-   `read.csv()`: Check the different options using `?read.csv`\
-   `str()`: Structure of an object\
-   `table()`: Check the optional arguments!\
-   `merge()`: Combine to data.frames\
-   `as.character()`: Change a vector's class to character
-   `as.numeric()`: Change a vector's class to numeric.
-   `%in%`
-   `is.na()`\
-   `max()`
-   `summary()`\
-   `complete.cases()`\
-   `cor.test()`
-   `%>%` and `group_by()` and `summarize()` from the dlyr package (check demonstration Part 2)

Regarding the solutions, note we don't expect you to come up with exactly this code - there are many ways to solve the problem. The solutions just give you an idea of a possible solution

### 1. Read data

Read the provided datasets *treedata.csv* and *species.csv* to R. Use the option **stringsAsFactors = FALSE** in the function `read.csv`.

Read the dataset *treedata.csv* and call it treedata. It has the following columns:

-   `species_code`: tree species code\
-   `dbh`: diameter at breast height (1.3 m) \[cm\]\
-   `height`: total height of the tree \[m\]. Measured only on a subset of trees.\
-   `rem`: coded values, containing remarks

Read the dataset *species.csv* and call it species. The dataset consists of the following columns:

-   `species_code`: tree species code (matching the one used in treedata.csv)\
-   `species_scientific`: Scientific species name\
-   `species_english`: English species name

------------------------------------------------------------------------

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

First, you read in the file using `read.csv`. You have to specify the correct separator for your dataset, here this is ";".

```{r, eval = F}
treedata <- read.csv('Where/ever/you/put/it/treedata.csv', sep = ";")
species <- read.csv('Hopefully/in/the/same/folder/species.csv', sep = ";")
```
:::

### 2. Dataset properties

Have a look at the properties of the dataset:

-   Which classes do the columns have?\
-   Did you expect these classes?

------------------------------------------------------------------------

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

The `data.frame` `dat` contains 4 columns: species, dbh (diameter at breast height \[cm\]), height \[m\] and bem1, a remark.\
We expect the following formats:

| column  | format    |
|---------|-----------|
| species | character |
| dbh     | numeric   |
| height  | numeric   |
| rem     | factor    |

Using `str` we get an overview of the structure of a dataset:

```{r}
str(treedata)
```

Column `dbh` is a `character`, altough we would have expected this one to be class numeric. This indicates, that a letter or special characters are in that column (we do not want these to be in there at all!).
:::

### 3. Turn character to numeric

One column, which we expect to be numeric, is of class character. Find the value causing this column to be character, set it to `NA` and turn the column into class numeric.

Note that using 'is.numeric()' is not enough, if the column is a factor. This may be the case if you have used the option **stringAsFactor = T** in `read.csv` or an older version of R. Use a combination of 'as.character()' and 'as.numeric()' in that case.

------------------------------------------------------------------------

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

We suspect `dbh` to contain a character and we want to remove this. With the function 'table()', we can check all existing values in the column. There seems to be an 'X' in the data.

```{r}
table(treedata$dbh)
```

To automatically search for characters, we can check if `dbh` contains a character that is part of `LETTERS` (capital letters) or `letters`:

```{r}
treedata[treedata$dbh %in% LETTERS | treedata$dbh %in% letters,]
```

A more advanced option would be to use `grepl`. If we are using the solution above, we will only find the value if it is exactly one character. Things get a bit more complicated, if we have special characters, e.g, a \*.

```{r}

x <- rep(c(1, 3, 5, '*', 'AA', ',', 9), 2)
x[grepl("^[A-Za-z]+$", x, perl = T)]
x[!grepl('[^[:punct:]]', x, perl =T)]

```

We want to set the X in `dbh` to `NA` (probably, this is a transcription error, so one could also have a look at the field forms...).

```{r}

treedata$dbh[treedata$dbh == 'X'] <- NA
str(treedata)

```

Just removing the 'X' does not turn a character to numeric! R provides the function `as.numeric`, which might be of use in this case.

```{r}
treedata$dbh <- as.numeric(treedata$dbh)
head(treedata$dbh)
```
:::

### 4. `NA`- values in `height`

Check for NA's in the column `height`:

-   How many `NA`'s do appear in this column?\
-   Did you expect this column to contain `NA`'s? Why?

------------------------------------------------------------------------

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

```{r}
summary(treedata)

sum(is.na(treedata$height))
table(is.na(treedata$height))

nrow(treedata[is.na(treedata$height),])
table(treedata$height, useNA = 'ifany')

```

-   As only on a subset of the trees, height has been measured, we expect this column to contain `NA`-values.\
-   The dataset contains 221 `NA`.
:::

### 5. Implausible values

The dataset contains some implausible values (completely out of range!). Find and replace these values with `NA`.

------------------------------------------------------------------------

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

*What values are implausible? The dataset contains different species. A value which is plausible for species A might be implausible for species B. However, for now, we will not go into the details here.*

Do a visual check of the dataset

```{r}
boxplot(treedata$height, main = 'Height')
boxplot(treedata$dbh, main ='DBH')

max(treedata$dbh, na.rm = T)
max(treedata$height, na.rm = T)

treedata[treedata$dbh > 500 & !is.na(treedata$dbh),]
treedata[treedata$height > 50 & !is.na(treedata$height),]
```

There seems to be one outlier in both datasets which can be seen as implausible: No tree is more than 100 m of height and no tree has a diameter \> 20 m (These values can be considered implausible for trees in Switzerland).\
We will now remove these values from our dataset by setting them to `NA` - this might not always be the best option, there are also statistical models that can account for such errors!

```{r}
treedata$dbh[treedata$dbh > 500] <- NA
treedata$height[treedata$height > 50] <- NA

boxplot(treedata$dbh, main = 'DBH')
boxplot(treedata$height, main = 'Height')
```
:::

### 6. Add species names

Add the species names from the species dataset to the treedata dataset.\
*Hint: `?merge`*

------------------------------------------------------------------------

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

Use `merge` to add species names to the dataset treedata. For adding only one column, `match` is a helpful function.

```{r}
treedata = merge(treedata, species, by = "species_code")
head(treedata)

# treedata$species_english <- species$species_english[match(treedata$species_code, species$species_code)]

```
:::

### 7. Remove F4

The remark F4 indicates, that the dbh-measurments might be flawed. Remove these trees from treedata.

------------------------------------------------------------------------

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

```{r}
treedata <- treedata[!treedata$rem %in% 'F4',]
```
:::

### 8. Select trees that contain both dbh and height measurements

Subset treedata to trees where both height and dbh measurements were carried out. Call the result dbh_height.

------------------------------------------------------------------------

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

This can be done using `complete.cases` using

```{r}
dbh_height <- treedata[complete.cases(treedata$dbh, treedata$height),]
```
:::

### 9. Correlations?

Check for correlations in the dataset dbh_height. Use a test and plot dbh vs height.

------------------------------------------------------------------------

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

```{r}
cor.test(dbh_height$dbh
         , dbh_height$height)

plot(dbh_height$dbh
     , dbh_height$height
     , ylab = 'Height'
     , xlab = 'DBH')
```
:::

### 10. Calculate mean dbh per species.

For calculating summary statistics, the dplyr package is really helpful. It is part of the tidyverse environment, which was designed for data science. If you work with large and complex datasets or if you have to derive many new variables, I really recommend that you have a look at this. Also, the syntax for dplyr is quite intuitive.

For help on this, check the demonstration of Part 2 where we calculated summary statistics for groups using the dplyr package!

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
#### Solution

```{r}
library(dplyr)
treedata %>% 
  group_by(species_english) %>% 
  summarize(N = n(),
            meanDBH = mean(dbh, na.rm = T), 
            sdDBH = sd(dbh, na.rm = T))
```
:::
